Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-11 benchmark-pca.Rmd
save

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # devtools::install_github("gabraham/flashpca", subdir = "flashpcaR")
  library(flashpcaR)
  # install.packages("tidyverse")
  library(tidyverse)
  # install.packages("glue")
  library(glue)   
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"

@@

@@

```{r}
# Preparing a dataset for the results
models <- tribble(
  ~method,               ~fun,     ~params,
  "FastPCA/EIGENSOFT",   fastpca,  list(),
  "FlashPCA2",           flashpca, list(),
  "bigstatsr - 1 core",  mypca,    list(),
  "bigstatsr - 6 cores", mypca,    list(ncores = NCORES)
) %>%
  mutate(size = list(c(1, 5, 10))) %>%
  unnest(size, .drop = FALSE) %>%
  mutate(time = NA, pcs = list(NA))

@@

```{r}
# SmartPCA
smartpca <- fastpca("backingfiles/celiacQC_keep1", fast = FALSE)
true <- smartpca[[2]]
cat(sprintf("Standard PCA takes forever: %.2gh.", smartpca[[1]] / 3600))
```

```{r}
# Tidy and plot results
(preci <- models %>% 
  mutate(
    size = as.factor(size),
    cor = lapply(pcs, function(x) {
      N <- nrow(x) / nrow(true)
      if (N == 1) {
        diag(cor(x, true))
      } else {
        diag(cor(x[c(TRUE, rep(FALSE, N - 1)), ], true))
      }
    }),
    PC = list(as.factor(1:10))
  ) %>%
  select(-c(fun, params, time, pcs)))

preci %>%
  unnest(cor, PC) %>%
  ggplot(aes(PC, cor^2, color = size, shape = size)) %>%
  bigstatsr:::MY_THEME(coeff = 2) +
  geom_point(cex = 5) +
  geom_line(aes(group = size, linetype = size), lwd = 2) +
  facet_wrap(~method, nrow = 2) +
  labs(x = "Principal Component", y = "Squared correlation between estimated and true PCs",
       color = "Number of individuals", shape = "Number of individuals", linetype = "Number of individuals") + 
  theme(legend.position = c(0.8, 0.15),
        strip.text.x = element_text(size = rel(3)),
        legend.key.width = unit(4.5, "line")) + 
  coord_cartesian(ylim = c(0, 1))

@@

```{r, include=FALSE}
library(gmailr)
send_message(
  mime(To = "florian.prive.21@gmail.com",
       From = "florian.prive.21@gmail.com",
       Subject = sprintf("Knitting of %s has finished",
                         knitr::current_input()))
)
```

>>>>

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # devtools::install_github("gabraham/flashpca/flashpcaR")
  library(flashpcaR)
  # install.packages("tidyverse")
  library(tidyverse)
  # install.packages("glue")
  library(glue)   
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"

# PLINK 2.0: https://www.cog-genomics.org/plink/2.0/
plink2 <- "./plink2"

@@

plink2pca <- function(prefix, fast = TRUE, parallel = TRUE) {

  time <- system.time(
    system(glue("{plink2} --bfile {prefix} --out {prefix}",
                " --pca {`if`(fast, 'approx 10', '10')}",
                "{`if`(!parallel, ' --threads 1', '')}"))
  )

  evec <- data.table::fread(glue("{prefix}.eigenvec"), 
                            select = paste0("PC", 1:10))

  list(time[3], as.matrix(evec))
}

@@

```{r}
# Preparing a dataset for the results
models <- tribble(
  ~method,               ~fun,       ~params,
  "FastPCA/EIGENSOFT",    fastpca,   list(),
  "FlashPCA2",            flashpca,  list(),
  "bigstatsr - 1 core",   mypca,     list(),
  "bigstatsr - 6 cores",  mypca,     list(ncores = NCORES),
  "PLINK 2.0 - 1 core",   plink2pca, list(parallel = FALSE),
  "PLINK 2.0 - 11 cores", plink2pca, list()
) %>%
  mutate(size = list(c(1, 5, 10))) %>%
  unnest(size, .drop = FALSE) %>%
  mutate(time = NA, pcs = list(NA))

@@

```{r}
# Exact PCA
true <- plink2pca("backingfiles/celiacQC_keep1", fast = FALSE)[[2]]
```

```{r}
# Tidy and plot results
(preci <- models %>%
    mutate(method = forcats::fct_recode(
      method, 
      `bigstatsr` = "bigstatsr - 1 core",
      `PLINK 2.0` = "PLINK 2.0 - 1 core", 
      NULL = "bigstatsr - 6 cores", 
      NULL = "PLINK 2.0 - 11 cores"
    )) %>%
    filter(!is.na(method)) %>%
    mutate(
      size = as.factor(size),
      cor = lapply(pcs, function(x) {
        N <- nrow(x) / nrow(true)
        if (N == 1) {
          diag(cor(x, true))
        } else {
          diag(cor(x[c(TRUE, rep(FALSE, N - 1)), ], true))
        }
      }),
      PC = list(as.factor(1:10))
    ) %>%
    select(-c(fun, params, time, pcs)))

preci %>%
  unnest(cor, PC) %>%
  ggplot(aes(PC, cor^2, color = size, shape = size)) %>%
  bigstatsr:::MY_THEME(coeff = 2) +
  geom_point(cex = 5) +
  geom_line(aes(group = size, linetype = size), lwd = 2) +
  facet_wrap(~method, nrow = 2) +
  labs(x = "Principal Component", y = "Squared correlation between estimated and true PCs",
       color = "Number of individuals", shape = "Number of individuals", linetype = "Number of individuals") + 
  theme(legend.position = c(0.25, 0.15),
        strip.text.x = element_text(size = rel(3)),
        legend.key.width = unit(4.5, "line")) + 
  coord_cartesian(ylim = c(0, 1))

@@

```{r, include=FALSE, error=TRUE}
library(gmailr)
send_message(
  mime(To = "florian.prive.21@gmail.com",
       From = "florian.prive.21@gmail.com",
       Subject = sprintf("Knitting of %s has finished",
                         knitr::current_input()))
)
```

### 2018-01-11 cabios-template2/bigsnpr-review.tex
save

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
For instance, one might prefer using R because it makes easier to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.

>>>>

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
For instance, one might like using R because R makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.
